Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations in the Upper Salmon River, Idaho

### 2021-12-15 analysis/paper/paper.Rmd
minor edits from Google doc

---
title: "Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations"
author:
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [mhe_mccall, biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Tara E. Blackman:
      email: tara.blackman@mthoodenvironmental.com
      institute: [mhe_sandy]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - mhe_mccall: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Fish Program, Science Division, 1111 Washington Street NE, Olympia, Washington, 98501, USA
  - mhe_sandy: Mount Hood Environmental, 39085 Pioneer Boulevard \#100 Mezzanine, Sandy, Oregon, 97055, USA 
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua      
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Abstract text here. 
keywords: |
  northern pikeminnow; Chinook salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

The Upper Salmon River major population group (MPG) supports eight independent, extant spring/summer Chinook Salmon *Oncorhynchus tshawytscha* populations, each providing important ecological, social, and economic benefits to the region. The distinct populations include Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017], all of which have become depleted in recent decades. Reduction in overall population size, and more specifically, declines in survival of juvenile Chinook Salmon have been associated with hydrosystem passage [@Schaller2014], ocean conditions, regional climate [@Welch2021; @Gosselin2021], and downstream predation by native and non-native piscivorous fishes [@Poe1991; @Friesen1999; @Harvey2004]. Anthropogenic changes to freshwater habitat including removal of beavers, river channel simplification, water withdrawals, and the proliferation of non-native species (e.g., non-native coastal rainbow trout *O. mykiss irideus* and brook trout *Salvelinus fontinalis*) have also been attributed to reductions in population size [@IdahoOSCTeam2019]. Moreover, the abundance of returning adults is further impacted by ocean and mainstem harvests, poor ocean conditions, and changes to the spawning migration corridor [@FCRPSBiOp2020]. Each of these factors have contributed, to varying extents, to reduced adult escapement, the primary metric used to assess population viability. In response to the decline in Upper Salmon Chinook Salmon abundance from the myriad of human activities and associated habitat degradation, action agencies have attempted to improve juvenile survival and adult spawning conditions by investing in rehabilitation of tributary ecosystems as well as investing in the study and evaluation of many factors attributed to the species decline [@Roni2018].

One potentially important, but perhaps under-appreciated source of mortality for Upper Salmon River Chinook Salmon is predation on emigrating juveniles by piscivorous fishes. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of population recovery, most notably in the presence of non-native piscivores [@Fresh2003; @Winther2019]. However, these studies have focused on large rivers and reservoirs of the Snake and Columbia rivers whereas very little is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems of the Columbia and Snake river [@Rubenson2020], including the Salmon River. Adding to this, there is little consensus within the literature related to the magnitude of predation on salmonids in less altered, free flowing systems, with some suggesting low levels of predation [@Brown1981; @Buchanan1981] while others suggest more impactful levels of predation [@Tabor1993; @Ward1995; @Shively1996; @Zimmerman1999] that may hinder recovery.

Dams and reservoirs in the Columbia River basin are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration rates of juvenile Chinook Salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces resulting from dams on the Columbia and Snake rivers have created favorable slow-water habitat for several fishes known to consume juvenile Chinook Salmon including both non-native (e.g., Smallmouth Bass *Micropterus dolomieu*, Walleye *Sander vitreus*) and native (e.g., Northern Pikeminnow *Ptychocheilus oregonensis*) species. In the Lower Granite Reservoir on the Snake River, juvenile Chinook Salmon are common prey items of Smallmouth Bass during some seasons [@Erhardt2018]. Similarly, non-native Walleye are considered a potentially important source for juvenile Chinook Salmon mortality in the Snake River [@Widener2021], particularly as their abundance has significantly increased surrounding Snake River dam facilities between 2008 and 2020.

@@

The slough is also part of the migratory pathway for the endangered Snake River Sockeye Salmon population and several populations of threatened Snake River steelhead. Accordingly, the mainstem Salmon River has been the target of recent radio telemetry studies to examine the downstream movement, distribution, and (apparent survival) of juvenile salmonid emigrants. For example, @Axel2015 examined the movement and survival of Sockeye Salmon released in Redfish Lake Creek downstream through the mainstem Salmon River during the spring migration and identified a substantial decrease in estimated survival in the reach spanning the Deadwater Slough. They estimated survival among release groups to be approximately 56% whereas survival through the majority of study reaches was greater than 80%. Further, recent telemetry studies of Chinook Salmon emigration during the fall and early winter months indicated a decreases transition probability in the reach spanning the Deadwater Slough, approximately 10% less than surrounding reaches [@Ackerman2018; @Porter2019].

@@

Multiple capture methods were employed during the fall 2019 effort to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. Unfortunately, some methods proved ineffective during initial efforts. For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. Eventually, angling proved to be the most effective method for capturing piscine predators while minimizing potential impacts to steelhead but recognizing that angling can have both species and size selectivity (bias). From here on, we focus on fish captured by angling, unless otherwise noted.

Our study relied heavily on volunteer anglers who were permitted to fish anywhere within Deadwater Slough. During a sample event, anglers would boat or hike their catch (periodically or upon filling a livewell) to a processing station located at the boat ramp approximately 500 m downstream from the top of the slough. For each fish, we recorded the date of capture, species, total length (TL; mm), and whether the fish was previously marked. Unmarked fish were given a unique physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) each day and released. Released individuals were distributed throughout Deadwater Slough to facilitate mixing back into the population. Finally, we recorded the angling start and end times for each crew (person or combinations of persons) to facilitate calculations of catch per unit effort (CPUE).

@@

We used several mark-recapture abundance estimators that each rely on meeting several assumptions:

1. The population is closed (no immigration, emigration, births or deaths during the sampling period)
1. Marking a fish does not affect its chances of being caught
1. No loss of marks
1. No marks are missed or mistaken
1. All fish have the same chance of being caught in the second (and subsequent) sample(s)

@@

The Lincoln-Petersen estimator can be biased with small samples, so we also investigated the Chapman-modified Lincoln-Petersen estimator:

@@

For a multiple census estimator we first considered the Schnabel estimator where the $M$, $n$ and $m$ are indexed by the sampling occassion, $i$. In our case, the sampling occassions are defined as each day of sampling. The Schnabel estimator does not have an associated standard error; however, the 90% confidence intervals can be calculated. The Schnabel estimator is essentially a weighted average of a series of Lincoln-Peterson estimators (with a Chapman modfication). The Schnabel estimator is shown here:

@@

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and fall 2020 sampling periods. However, due to logistical constraints, we were only able to sample during a single week in spring 2021 and released fish were left unmarked. Instead, to estimate the abundance of Northern Pikeminnow during spring 2021, we used the ratio of total CPUE in the fall sampling events to the total CPUE in the spring sampling event, and multiplied that ratio by the average abundance of Northern Pikeminnow from the fall sampling events.<!-- Kevin, is this previous statement accurate? --> This approach assumes equal capture probabilities between the fall and spring capture events. If that assumption is not true and capture probability did differ, we believe that the capture probability  in the spring would be lower due to higher flows, which would lead to our spring abundance estimate to be biased low. Therefore, we presume our spring abundance estimate during the NRR Chinook Salmon emigration to be a conservative estimate of the number of Northern Pikeminnow present in the Deadwater Slough.

@@

Lastly, we calculated the proportional stock density (PSD) for Northern Pikeminnow captured during all field efforts, which is a measure of species size structure. PSD is a ratio, typically expressed as a percentage, between the number of "quality-sized" individuals or larger individuals and "stock-sized" individuals. Stock and quality size definitions vary by species. PSD is expressed as:

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes (e.g., juvenile steelhead, juvenile Sockeye Salmon, Redside Shiner *Richardsonius balteatus*, etc.). Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks for later analysis in a controlled environment. For each sample, wet weight (grams) was recorded for the total combined stomach content, including all non-fish items (e.g., macroinvertebrates, organic matter), as well as the portion consisting of fish parts/matter. Fish and fish remnants were identified to the lowest taxonomic unit, when possible, or were categorized as unknonw. A subset of Northern Pikeminnow captures (~ 5%) were euthanized for dissection after gastric lavage to validate the efficacy of the methodology.

@@

To estimate the total consumption potential (i.e., number/amount of juvenile consumed during the peaks of DSR [fall] and NRR [spring] emigrations) of Northern Pikeminnow in Deadwater Slough, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 applied in the R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook Salmon based on an estimate from @Moss2016. We used a weight-length formula [@Parker1995] to calculate the average predator start weight from the average length of Northern Pikeminnow captured during our study using the FSA package [@Ogle2021] in R. Continuous water temperature data from 2019 was used in two alternative models, one using the baseline start and end weights and the second with a 10% increase in average end weight on Northern Pikeminnow.

@@

To estimate the total biomass consumed, we then multipled biomass consumed by an individual Northern Pikeminnow for the given time period by the estimated population size for that time period. The total biomass was further converted to an estimate of the total number of Chinook Salmon consumed by dividing by a weight of 12 g per individual. This weight was an average from juvenile Chinook Salmon emigrating past six rotary screw traps operating in the Upper Salmon above Deadwater Slough including traps in the Lemhi, Pahsimeroi, and North Fork Salmon rivers and one trap operating near the Sawtooth hatchery.

@@

We caught `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "n_fish"])` and `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "n_fish"])` Northern Pikeminnow during Fall 2019 and Fall 2020, respectively. The catch per unit effort, measured as number of Northern Pikeminnow captured per angler hour, for each was `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "catch_per_hr"], digits = 3)` for Fall 2019 and `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "catch_per_hr"], digits = 3)` for Fall 2020 (Figure \@ref(fig:cpue-fig)). During Fall 2019 we had a total of `r as.english(sum(N_mods[[3]][[1]]$m))` recaptures which included `r as.english(LP_mod[[3]][[1]]$m)` unique individuals i.e. `r as.english(sum(N_mods[[3]][[1]]$m) - LP_mod[[3]][[1]]$m)` was captured more than once. There were `r as.english(sum(N_mods[[3]][[3]]$m))` recaptures in Fall 2020 which included `r as.english(LP_mod[[3]][[3]]$m)` unique individuals. Summaries of parameters for both the single-census and multiple-census mark-recapture estimators are provided in Table \@ref(tab:abund-data-lp) and Table \@ref(tab:abund-data-schn), respectively.

@@

We captured an additional `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "n_fish"])` Northern Pikeminnow during Spring 2021 and CPUE was `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "catch_per_hr"], digits = 2)` fish per angler hour. ***Finish spring abundance results.*** <!-- Can we assume equal capture probability and perhaps using the ratio of the average fall CPUE and spring CPUE. Then multiply the average fall abundance estimate by the ratio. Does that seem appropriate? -->

>>>>

---
title: "Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations in the Upper Salmon River, Idaho"
author:
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [mhe_mccall, biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Tara E. Blackman:
      email: tara.blackman@mthoodenvironmental.com
      institute: [mhe_sandy]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - mhe_mccall: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Fish Program, Science Division, 1111 Washington Street NE, Olympia, Washington, 98501, USA
  - mhe_sandy: Mount Hood Environmental, 39085 Pioneer Boulevard \#100 Mezzanine, Sandy, Oregon, 97055, USA 
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua      
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Abstract text here. 
keywords: |
  northern pikeminnow; Chinook salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

The Upper Salmon River major population group (MPG) supports eight independent, extant spring/summer Chinook Salmon *Oncorhynchus tshawytscha* populations, each providing important ecological, social, and economic benefits to the region. The distinct populations include Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017], all of which have become depleted in recent decades. Reduction in overall population size, and more specifically, declines in survival of juvenile Chinook Salmon have been associated with hydrosystem passage [@Schaller2014], ocean conditions, regional climate [@Welch2021; @Gosselin2021], and downstream predation by native and non-native piscivorous fishes [@Poe1991; @Friesen1999; @Harvey2004]. Anthropogenic changes to freshwater habitat including removal of beavers, river channel simplification, water withdrawals, and the proliferation of non-native species (e.g., non-native coastal rainbow trout *O. mykiss irideus* and brook trout *Salvelinus fontinalis*) have also been attributed to reductions in population size [@IdahoOSCTeam2019]. Moreover, the abundance of returning adults is impacted by ocean and mainstem harvests, poor ocean conditions, and changes to the spawning migration corridor [@FCRPSBiOp2020]. Each of these factors have contributed, to varying extents, to reduced adult escapement, the primary metric used to assess population viability. In response to the decline in Upper Salmon Chinook Salmon abundance from the myriad of human activities and associated habitat degradation, action agencies have attempted to improve juvenile survival and adult spawning conditions by investing in rehabilitation of tributary ecosystems as well as investing in the study and evaluation of many factors attributed to the species decline [@Roni2018].

One potentially important, but perhaps under-appreciated source of mortality for Upper Salmon River Chinook Salmon is predation on emigrating juveniles by piscivorous fishes. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of population recovery, most notably in the presence of non-native piscivores [@Fresh2003; @Winther2019]. However, these studies have focused on large rivers and reservoirs of the Snake and Columbia rivers whereas very little is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems of the Columbia and Snake river [@Rubenson2020], including the Salmon River. Adding to this, there is little consensus within the literature about the magnitude of predation on salmonids in less altered, free flowing systems, with some suggesting lower rates of predation [@Brown1981; @Buchanan1981] while others suggest more impactful rates of predation [@Tabor1993; @Ward1995; @Shively1996; @Zimmerman1999] that may hinder recovery.

Dams and reservoirs in the Columbia River basin are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, movement (swimming) rates of juvenile Chinook Salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces resulting from dams on the Columbia and Snake rivers have created favorable slow-water habitat for several fishes known to consume juvenile Chinook Salmon including both non-native (e.g., Smallmouth Bass *Micropterus dolomieu*, Walleye *Sander vitreus*) and native (e.g., Northern Pikeminnow *Ptychocheilus oregonensis*) species. In Lower Granite Reservoir on the Snake River, juvenile Chinook Salmon are common prey items of Smallmouth Bass during some seasons [@Erhardt2018]. Similarly, non-native Walleye are considered a potentially important source for juvenile Chinook Salmon mortality in the Snake River [@Widener2021], particularly as their abundance has significantly increased surrounding Snake River dam facilities between 2008 and 2020.

@@

The slough is also part of the migratory pathway for the endangered Snake River Sockeye Salmon population [@Axel2015] and several populations of threatened Snake River steelhead. Accordingly, the mainstem Salmon River has been the target of recent radio telemetry studies to examine the downstream movement, distribution, and (apparent survival) of juvenile salmonid emigrants. For example, @Axel2015 examined the movement and survival of Sockeye Salmon released in Redfish Lake Creek downstream through the mainstem Salmon River during the spring migration and identified a substantial decrease in estimated survival in the reach spanning the Deadwater Slough. They estimated survival among release groups to be approximately 56% whereas survival through the majority of study reaches was greater than 80%. Further, recent telemetry studies of Chinook Salmon emigration during the fall and early winter months indicated a lower transition probability in the reach spanning the Deadwater Slough, approximately 10% less than surrounding reaches [@Ackerman2018; @Porter2019].

@@

Multiple capture methods were employed during the fall 2019 effort to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. However, some methods proved ineffective during initial efforts. For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. After evaluating all methods, angling proved to be the most effective method for capturing piscine predators while minimizing potential impacts to steelhead but recognizing that angling can have both species and size selectivity (bias). From here on, we focus on fish captured by angling, unless otherwise noted.

Our study relied heavily on volunteer anglers who were permitted to fish anywhere within Deadwater Slough. During a sample event, anglers would boat or hike their catch (periodically or upon filling a livewell) to a processing station located at the boat ramp approximately 500 m downstream from the top of the slough. For each fish, we recorded the date of capture, species, total length (TL; mm), and whether the fish was previously marked. Unmarked fish were given a physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) unique to each day and released. Released individuals were distributed throughout Deadwater Slough to facilitate mixing back into the population. Finally, we recorded the angling start and end times for each crew (person or combinations of persons) to facilitate calculations of catch per unit effort (CPUE).

@@

All mark-recapture abundance estimators assumed: (1) the population is closed (no immigration, emigration, births or deaths during the sampling period), (2) marking a fish does not affect its chances of recapture, (3) no loss of marks, (4) marks are not missed or mistaken, and (5) all fish have equal chance of being caught in the second (and subsequent) sample(s).

@@

The Lincoln-Petersen estimator can be biased with small samples, so we also employed the Chapman-modified Lincoln-Petersen estimator:

@@

For a multiple-census estimator, we first utilized the Schnabel estimator where the $M$, $n$ and $m$ are indexed by the sampling occasion, $i$. In our case, the sampling occassions are defined as each day of sampling. The Schnabel estimator does not have an associated standard error; however, the 90% confidence intervals can be calculated. The Schnabel estimator is essentially a weighted average of a series of Lincoln-Peterson estimators (with a Chapman modfication). The Schnabel estimator is shown here:

@@

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and fall 2020 sampling periods. However, due to logistical constraints, we were only able to sample during a single week in spring 2021 and released fish were left unmarked. Instead, to estimate the abundance of Northern Pikeminnow during spring 2021, we used the ratio of total CPUE in the fall sampling events to the total CPUE in the spring sampling event, and multiplied that ratio by the average abundance of Northern Pikeminnow from the fall sampling events.<!-- Kevin, is this previous statement accurate? --> This approach assumes equal capture probabilities between the fall and spring capture events. If that assumption is not true and capture probability did differ, we believe that the capture probability  in the spring would be lower due to higher flows, which would lead to our spring abundance estimate to be biased low. Therefore, we presume our abundance estimate of Northern Pikeminnow for the spring 2021 sampling event to be a conservative estimate.

@@

Finally, we calculated the proportional stock density (PSD) for Northern Pikeminnow captured during all field efforts, which is a measure of species size structure. PSD is a ratio, typically expressed as a percentage, between the number of "quality-sized" individuals or larger individuals and "stock-sized" individuals. Stock and quality size definitions vary by species. PSD is expressed as:

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes (e.g., juvenile steelhead, juvenile Sockeye Salmon, Redside Shiner *Richardsonius balteatus*, etc.). Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks for later analysis in a controlled environment. For each sample, wet weight (grams) was recorded for the total combined stomach content, including all non-fish items (e.g., macroinvertebrates, organic matter), as well as the portion consisting of fish parts/matter. Fish and fish remnants were identified to the lowest taxonomic unit, when possible, or were categorized as unknown. A subset of Northern Pikeminnow captures (~ 5%) were euthanized for dissection after gastric lavage to validate the efficacy of the methodology.

@@

To estimate the total consumption potential (i.e., number/amount of juvenile consumed during the peaks of DSR [fall] and NRR [spring] emigrations) of Northern Pikeminnow in Deadwater Slough, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 applied in the R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook Salmon based on an estimate from @Moss2016. We used a weight-length formula [@Parker1995] to calculate the average predator start weight from the average length of Northern Pikeminnow captured during our study using the FSA package in R [@Ogle2021]. The average length of Northern Pikeminnow captured at the Deadwater Slough during the spring and fall were 352.9 and 393.7 mm, respectively, which calculates to average starting weights of 504.4 and 670.5 g. Continuous water temperature data from 2019 was used in two alternative models, one using the baseline start and end weights and the second with a 10% increase in average end weight on Northern Pikeminnow.

@@

To estimate the total biomass consumed, we then multiplied the biomass consumed by an individual Northern Pikeminnow for the given time period by the estimated population size for that time period. The total biomass was further converted to an estimate of the total number of Chinook Salmon consumed by dividing by a weight of 12 g per individual. This weight was an average from juvenile Chinook Salmon emigrating past six rotary screw traps operating in the Upper Salmon above Deadwater Slough including traps in the Lemhi, Pahsimeroi, and North Fork Salmon rivers and one trap operating near the Sawtooth hatchery.

@@

We caught `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "n_fish"])`, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "n_fish"])`, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "n_fish"])` Northern Pikeminnow during Fall 2019, Fall 2020, and Spring 2021, respectively. The catch per unit effort, measured as the number of Northern Pikeminnow captured per angler hour, for each was `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "catch_per_hr"], digits = 3)` for Fall 2019, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "catch_per_hr"], digits = 3)` for Fall 2020, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "catch_per_hr"], digits = 2)` for Spring 2021 (Figure \@ref(fig:cpue-fig)). During Fall 2019, we had a total of `r as.english(sum(N_mods[[3]][[1]]$m))` recaptures which included `r as.english(LP_mod[[3]][[1]]$m)` unique individuals (i.e., `r as.english(sum(N_mods[[3]][[1]]$m) - LP_mod[[3]][[1]]$m)` individual was captured more than once). There were `r as.english(sum(N_mods[[3]][[3]]$m))` recaptures in Fall 2020 which included `r as.english(LP_mod[[3]][[3]]$m)` unique individuals. Summaries of parameters for both the single-census and multiple-census mark-recapture estimators are provided in Table \@ref(tab:abund-data-lp) and Table \@ref(tab:abund-data-schn), respectively.

@@

***Finish spring abundance results.*** <!-- Can we assume equal capture probability and perhaps using the ratio of the average fall CPUE and spring CPUE. Then multiply the average fall abundance estimate by the ratio. Does that seem appropriate? -->
